Safe Transferable Regions

### 2015-11-19 paper/abstract.tex
Edits.

Manual memory management is susceptible to the usual perils of
dangling pointers.  A key contribution of this paper is a region type
system that ensures the memory safety of C\# programs in the presence
of transferable regions. We compliment our type system with a type
inference algorithm that infers principal region types for first-order
programs. This eliminates the need for programmers to write region
annotations on types, while facilitating the reuse of existing C\#
libraries with no modifications. Various case studies demonstrate the
practical utility of our approach.

>>>>

Manual memory management is susceptible to the usual perils of
dangling pointers.  A key contribution of this paper is a region type
system that ensures the memory safety of C\# programs in the presence
of transferable regions. We compliment our type system with a type
inference algorithm that infers principal region types for first-order
programs, and useful types for higher-order programs. This eliminates
the need for programmers to write region annotations on types, while
facilitating the reuse of existing C\# libraries with no
modifications. Experiments demonstrate the practical utility of our
approach.

### 2015-11-19 paper/related.tex
Edits.

~\cite{HMN01} proposes a flow-sensitive region-based memory management
for first-order programs that proposes to overcome some of the
drawbacks of~\cite{tofte97} by generalizing~\cite{tofte97}'s approach
to regions with dynamic lifetimes. However, dynamic regions are still
not first-class values of the language, and reference counting is
nonetheless needed to ensure memory safety.~\cite{WW01} extends lambda
calculus with first-class regions with dynamic lifetimes, and imposes
linear typing to control accesses to regions. All type annotations
need to be written explicitly. Moreover, invariants needed to ensure
safety under region transfers, such as the absence of references that
escape a transferable region, cannot be enforced under~\cite{WW01}'s
type system.

>>>>

~\cite{HMN01} proposes a flow-sensitive region-based memory management
for first-order programs that proposes to overcome some of the
drawbacks of~\cite{tofte97} by generalizing~\cite{tofte97}'s approach
to regions with dynamic lifetimes. However, dynamic regions are still
not first-class values of the language, and reference counting is
nonetheless needed to ensure memory safety.~\cite{WW01} extends lambda
calculus with first-class regions with dynamic lifetimes, and imposes
linear typing to control accesses to regions. Our open/close lexical
block for transferable regions traces its orgins to the \C{let!}
expression in~\cite{WW01} and~\cite{wadler90} that safely relaxes
linear typing restrictions, allowing variables to be temporarily
aliased. We don't have linear typing, thus admit unrestricted
aliasing. Non-linear typing keeps our type system simple, and makes it
possible to infer types, thus eliminating the annotation burden.
Moreover, ~\cite{WW01}'s linear type system is insufficient to enforce
the invariants needed to ensure safety under region transfers, such as
the absence of references that escape a transferable region.

### 2015-11-19 paper/type-inference.tex
Edits.

\subsection{Constraint Solving}
\label{sec:fb-constraintsolving}

@@

% The region type system for $\FB$ admits function types with
% higher-rank region polymorphism, allowing region variables to be
% quantified in the types of higher-order arguments. This allows, for
% example, a user-defined function to be applied on arguments from
% different regions in different contexts. However, like most type
% systems, our region type system loses principal typing property in
% presence of higher-rank polymorphism, which means that there can be
% two possible types for a function, where neither one is more general
% than the other. Consequently, it is not possible to infer a unique
% type for a function. Nonetheless, it is possible to engineer type
% inference to to consistently prefer one type over the other.
Our type inference algorithm does not generalize to higher-order case,
due to the lack of principal typing in presence of higher-rank
polymorphism. We therefore adopt a trivial approach of assigning the
same region to the arguments and return values in (higher-order)
function types. For user-defined functions (e.g., predicate in the
\C{WHERE} clause of the \C{SELECT} dataflow query) and lambda
arguments to the $\RgnZ$ class constructor, which are the prominent
uses of higher-order functions in \name, the simplistic approach seems
to suffice in practice. However, it is always possible that this
approach makes type checker reject a program that is otherwise valid.
In such cases, programmer can help the type checker by writing region
annotations for only the higher-order functions.

>>>>

\subsection{Constraint Solving}
\label{sec:csolve}

@@

% The region type system for $\FB$ admits function types with
% higher-rank region polymorphism, allowing region variables to be
% quantified in the types of higher-order arguments. This allows, for
% example, a user-defined function to be applied on arguments from
% different regions in different contexts. However, like most type
% systems, our region type system loses principal typing property in
% presence of higher-rank polymorphism, which means that there can be
% two possible types for a function, where neither one is more general
% than the other. Consequently, it is not possible to infer a unique
% type for a function. Nonetheless, it is possible to engineer type
% inference to to consistently prefer one type over the other.
Our type inference algorithm does not straightforwardly generalize to
the higher-order case, due to the lack of principal typing in presence of
higher-rank polymorphism. Fortunately, higher-order type inference, in
its full generality, is rarely needed in the context of \name, where
higher-order control flow is predominantly due to the second-order
arguments to constructors (For \eg, user-defined functions in dataflow
queries, and lambda arguments to the $\RgnZ$ class constructor). We
therefore extend our inference algorithm with a heuristics tailor-made
for the aforementioned second-order cases. The simplistic approach
seems to suffice in our experience, but, it is always possible that it
makes the type checker reject a program that is otherwise valid. In
such cases, programmer can help the type checker by writing region
annotations only over higher-order arguments.  Nonetheless, we note
that it is possible to infer most general (if not principal) types for
higher-order arguments, by extending our constraint solver with an
approach modeled after~\cite{gulwani09}, which lets it compute maximally
weak/strong solutions to multiple competing $\varphi$'s
simultaneously.
